Stability and Chaos in Multi-agent Reinforcement Learning

### 2020-11-24 aamas2021-formatting-instructions/Responses to Reviews for AAMAS.txt

Review 1:

We thank the reviewer for their comments.

FB: it is not entirely clear to me what are you referring to in the following paragraph. I would suggest to focus on the question for rebuttal and leave this for later. 
Regarding your point on the links to evolutionary dynamics, actually the paper by Tuyls' et al (2006) show that, by deriving the continuous time dynamics of Q-Learning, we arrive at a dynamical system which is similar to the replicator dynamic from Evolutionary Game Theory. Our goal in this paper is solely focused on understanding the stability of these dynamics.

As regards specifically the questions for rebuttal:

1. We aim to focus our study on all normal form games. As you have pointed out, whilst the strategy space of the agents includes a finite set of actions, they are able to play mixed strategies as well as pure ones. The goal of the paper is to identify the evolution of the strategy as the agents learn. Thank you for pointing this out, we will be sure to clarify it in revision.

2 and 3. These are excellent points and we are, in fact, using the results of this study to address the problem of how to allow the agents to 'steer' the learning towards stable equilibria. However, as you have mentioned, there are cases in which there are multiple equilibria. In such cases, we have not yet determined a closed form method for steering the system towards a given one. This is an incredibly important question, however, and an important avenue for future work.

Review 2:

Thanks for the insightful remarks. Indeed, many of the techniques applied in this study are inspired by statistical physics, including the analysis of dynamical systems. We believe that such methods can be fruitfully applied to learning on games (as already shown by a wealth of works in the literature), as these contains many random variable elements: the strategies of the agents change during play, but also the payoff elements themselves can vary from game to game. We aimed, as much as possible, to focus the paper on the main, theoretical points of our contribution, while providing as many details as possible about the derivation in the supplementary material. We thank you for taking the time to look at the supplementary material, and for doing so with such rigour.

As you mentioned, the main contribution of this paper is equation (13) which establishes the region of stability. To validate this equation, we were sure to test each of its predictions, including the rather remarkable ones that independent Q-Learning rarely converges, and that the payoff correlations make no difference to stability.

As regards the style of presentation, we are aware that normally, papers at AAMAS present results in a theorem-proof format. However, we felt that this format would not be appropriate for the results described herein, as the statement of each result would just amount to the last equation in the derivation.

Review 3:

We thank the reviewer for the several points they raise, they will certainly be useful to improve the paper. They also allow us to provide some clarification, even though the reviewer does not ask any question explicitly.

The reviewer mentions that there is a disconnect between the theoretical analysis and the experimental evaluation. Nonetheless, we find that the theoretical result (13) makes a number of predictions including: the decrease of stability for increasing alpha or tau, the (rather surprising) lack of dependence on Gamma (the correlation between payoff elements), as well as the fact that the majority of the phase space is unstable. These predictions hold for the limiting case of an infinite action set and a large number of players. However, there are practical restrictions when conducting an experiment with an incredibly large action/player set. Namely, this requires storage and calculations on matrices with N^p elements. This results in severe limitations for the scale of the experiments that can be run. We do however, present a number of results in our empirical evaluation which show that the overall trend, i.e., as p and N increase. Consider, Figure 6, which considers the case of 6 players. We show that the trend is towards the theoretical predictions. It is in this sense that we mention that the size of the unstable region is overestimated by the theory, a point which was also found in [15]. The message of the paper is to highlight the sensitivity of the stability to p and N, and argue that it is required for algorithms to be analysed with stability in mind, as well as to design algorithms which focus on maintaining the stability of the system.

One of your queries regards the sentence "... in the Prisoner's Dilemma, cyclic behaviour emerges". The paper mentioned in this context is actually different from that of Figure 1. In the paper you are referencing, the authors make use of an established dynamics of the 'Experience Weighted Attraction' Learning algorithm, which shows cyclic behaviour. However, Figure 1 depicts the convergence of Q-Learning in the same PD problem. We will ensure that we clarify this distinction since it shows that different algorithms may give different results on the same problem, hence the need for such a convergence study.

Regarding the class of games considered: In this study, we aimed to make as little assumptions on the game as possible. Therefore, you are correct in saying that the payoff elements were already in a general form. However, even within this, we have the different classes of games (zero-sum, potential and everything in between). Further still, there are an infinite number of different payoff elements which could achieve a zero-sum game. We aim to parameterise the structure of the game by the Gamma, which measures the 'zero-sumness' of the game. Therefore, our study is focused on how this parameter affects the stability of the system, rather than having to vary every payoff element. It is in this sense that we make the assertion that we generalise the nature of the study to its minimal elements: the learning parameters and a parameterisation of the payoff matrix.

Your point: The convergence criterion used in the empirical evaluation should depend on the step size alpha, otherwise for a given alpha, 1% might be outside of the margin of error.

Response: (for francesco) I'm not sure what they mean by this, the empirical evaluation is already based on alpha?

Regarding the choice of tau: You are right that the values chosen are rather small. We chose these for two reasons. The first was to remain consistent with the results provided in [15], thereby justifying the comparison to it made in the discussion. Further, as our results show, the degree of instability rises as alpha and tau increase. Therefore, it is immediate to see that, for higher values of tau and alpha, the game is trivially unstable and so does not show the transition to stability.

### 2020-11-24 aamas2021-formatting-instructions/Responses to Reviews for AAMAS.txt

We thank the reviewer for their comments. As regards specifically the questions for rebuttal:

1. We aim to focus our study on all normal form games. As you have pointed out, whilst the strategy space of the agents includes a finite set of actions, they are able to play mixed strategies as well as pure ones. The goal of the paper is to identify the evolution of the strategy as the agents learn. Thank you for pointing this out, we will be sure to clarify it in revision.

2 and 3. These are excellent points and we are, in fact, using the results of this study to address the problem of how to allow the agents to 'steer' the learning towards stable equilibria. However, as you have mentioned, there are cases in which there are multiple equilibria. In such cases, we have not yet determined a closed form method for steering the system towards a given one. This is an incredibly important question, however, and an important avenue for future work.

Finally, regarding your point on the links with evolutionary dynamics (weaknesses), we notice that [20] shows that, by deriving the continuous time dynamics of Q-Learning, we arrive at a dynamical system which is similar to the replicator dynamic from Evolutionary Game Theory. Our goal in this paper is solely focused on understanding the stability of these dynamics.

@@

As regards the style of presentation, we are aware that normally, papers at AAMAS present results in a theorem-proof format. However, we felt that this format would not be appropriate for the results described herein, as the statement of each result would just amount to the last equation in the derivation.

Review 3:

We thank the reviewer for the several points they raise, they will certainly be useful to improve the paper. They also allow us to provide some clarification, even though the reviewer does not ask any question explicitly.

The reviewer mentions that there is a disconnect between the theoretical analysis and the experimental evaluation. Nonetheless, we find that the theoretical result (13) makes a number of predictions including: the decrease of stability for increasing alpha or tau, the (rather surprising) lack of dependence on Gamma (the correlation between payoff elements), as well as the fact that the majority of the phase space is unstable. These predictions hold for the limiting case of an infinite action set and a large number of players. However, there are practical restrictions when conducting an experiment with a large action/player set. Namely, this requires storage and calculations on matrices with N^p elements.
As the reviewer remarks, this results in severe limitations for the scale of the experiments that can be run. We do however, present a number of results in our empirical evaluation which show that the overall trend, i.e., as p and N increase. Consider, Figure 6, which considers the case of 6 players. We show that the trend is towards the theoretical predictions. It is in this sense that we mention that the size of the unstable region is overestimated by the theory, a point which was also found in [15]. The message of the paper is to highlight the sensitivity of the stability to p and N, and argue that it is required for algorithms to be analysed with stability in mind, as well as to design algorithms which focus on maintaining the stability of the system.

### 2020-11-24 Documents/ChaoticDynamics/aamas2021-formatting-instructions/Responses to Reviews for AAMAS.txt

Review 1:

We thank the reviewer for their comments. For the questions:

1. We aim to focus our study on all normal form games. As you have pointed out, whilst the strategy space of the agents includes a finite set of actions, they are able to play mixed strategies as well as pure ones. The goal of the paper is to identify the evolution of the strategy as the agents learn. Thank you for pointing this out, we will be sure to clarify it in revision.

2 and 3. These are excellent points and we are, in fact, using the results of this study to consider how to 'steer' the learning towards stable equilibria. As you have mentioned, there are cases in which there are multiple stable equilibria. In such cases, we have not yet determined a closed form method for steering the system towards a given one. This is an incredibly important question, however, and an important avenue for future work.

Finally, on the links to evolutionary dynamics, (weaknesses), we notice that [20] shows that, by deriving the continuous time dynamics of Q-Learning, we arrive at a system which similar to the replicator dynamic . Our goal in this paper is solely focused on understanding the stability of these dynamics.

Review 2:

Thanks for the insightful remarks. Indeed, many of the techniques applied in this study are inspired by statistical physics. We believe that such methods can be fruitfully applied to learning on games (as already shown by a wealth of works in the literature), as these contains many random variable elements: the strategies of the agents change during play, but also the payoff elements themselves can vary from game to game. We aimed, as much as possible, to focus the paper on the main, theoretical points of our contribution, while providing all details about the derivation in the supplementary material. We thank you for taking the time to look at the supplementary material with such rigour.

As you mentioned, the main contribution of this paper is equation (13) which establishes the region of stability. To validate this equation, we were sure to test each of its predictions, including the rather remarkable ones that independent Q-Learning rarely converges, and that the payoff correlations make no difference to stability.

As regards the style of presentation, we are aware that normally, papers at AAMAS present results in a theorem-proof format. However, we felt that this format would not be appropriate for the results described herein, as the statement of each result would just amount to the last equation in the derivation.

Review 3:

We thank the reviewer for the several points they raise, they will certainly be useful to improve the paper, and also to provide some clarification.

The reviewer mentions that there is a disconnect between the theoretical analysis and the experimental evaluation. Nonetheless, we find that the theoretical result (13) makes a number of predictions including: the decrease of stability for increasing alpha or tau, the (rather surprising) lack of dependence on Gamma (the correlation between payoff elements), and, for the limiting case of an infinite action set and a large number of players, that the majority of the phase space is unstable. There are, however, practical restrictions when conducting an experiment with an incredibly large action/player set; this requires storage and calculations on matrices with N^p elements, resulting in severe limitations for the scale of the experiments that can be run. We do however, present a number of results in our empirical evaluation which show the overall trend, i.e., as p and N increase. Consider, Figure 6, which considers 2-6 players, and shows the trend is towards the theoretical predictions. It is in this sense that the size of the unstable region is overestimated by the theory, a point which was also found in [15]. The message of the paper is to highlight the dependence of stability to these parameters, and argue that it is required for algorithms to be analysed with stability in mind, and to design algorithms which focus on maintaining stability.

One of your queries regards the sentence "... in the Prisoner's Dilemma, cyclic behaviour emerges". The study this references is actually different to that which Figure 1 depicts and, in fact, discusses a different learning algorithm (EWA) to that of the figure (Q-Learning). We will ensure that we clarify this distinction.

Method:

As discussed in the preceding paragraph to (4), i_mu refers to the action taken by player mu whilst i_kappa refers to that taken by kappa. This allows us to forego using different letters (i,j,k...) for different players, when we do not assume how many players there are.

The point regarding scaling the system was given in brief on page 4 to conserve space. A detailed discussion can be found in Section S2.2 of the supplementary material.

You are correct in saying that the payoff elements were already in a general form. To then analyse the system we parameterised these payoffs by 'Gamma' which measures the correlation between elements. To study the expected behaviour of a system which would result from a given Gamma, we average over all the possible realisations that this choice could take. This means that we can focus on how, on average, the payoff correlations affect stability rather than look at all of the (infinite) possible variations of payoff matrices.

Experiment:

The value of tau was chosen for two reasons 
1) to remain consistent with the results provided in [15], thereby justifying the comparison made in the discussion. 
2) as our results show, the degree of instability rises with tau. It is immediate to see that, for higher values of tau, the game is trivially unstable everywhere.

Your reading of the 'variance of actions' is correct, it denotes how much the actions vary about the fixed point. We will be sure to explicitly state how x_i(t) should be read

TODO: Make the points in the first paragraph, PD and Game assumptions as brief as possible. Respond to comments on notation, scaling the system and variance of the actions

### 2020-12-03 Documents/ChaoticDynamics/aamas2021-formatting-instructions/aamas21-formatting-instructions/sample.tex
Updated AAMAS Submission

\begin{abstract}
Modelling the dynamics of Q-Learning is an active and important topic
for the sake of developing an {\em a priori} understanding of
Reinforcement Learning. In this paper, we use methods from
evolutionary game theory to analyse the stability of Q-Learning in
$p$-player games, where payoffs are randomly generated. We determine
the parameter range in which Q-Learning is expected to settle to a
stable fixed point and the range in which the dynamics are unstable.

This study allows for parameters to be appropriately chosen to ensure
the safe convergence of a learning algorithm and as a first step
towards understanding the range of behaviours that can be displayed by
learning using the Q-Learning algorithm. We validate our theoretical
results through numerical simulation and show that, within the bounds
of experimental error, the region of instability can be characterised
by the learning dynamics.
\end{abstract}

@@

Single agent reinforcement learning (RL) is a well-established
framework for allowing agents to learn optimal strategies when trained
on an iterated task
%For a particularly strong example of this see
\cite{Vinyals2019}. However, for the realisation of complex tasks,
such as air traffic control, market negotiations, and multi-robot
coordination, it is required that the system be modelled as a
multi-agent system (MAS). Such systems are not as well-understood as
single-agent systems in RL, since a given agent is tasked with
optimising a reward function which depends not only on a
non-stationary environment, but also on the actions of other, possibly
loosely coupled, agents \cite{SchwartzMulti-agentApproach}.

It is, therefore, of paramount importance to develop a strong
theoretical understanding of multi-agent reinforcement learning (MARL)
to allow an {\em a priori} understanding of the behaviour of a given
learning algorithm. Fortunately, the study of MAS is not unique to the
field of AI and has been extensively investigated from the point of view of
economics and game theory as well
\cite{ShohamMultiagentFoundations}. In particular,
%the study of
evolutionary game theory (EGT) considers the problem of a MAS which is
repeatedly exposed to an iterated game. This idea shares a strong
resemblance with MARL and, in fact, in \cite{Tuyls2006AnGames} it was
shown that techniques from EGT may be fruitfully applied to the analysis of
Q-Learning \cite{Sutton2018}.

@@

\paragraph{Contribution}
In this study, we will be considering the question: under what choice
of parameters is a multi-agent system, which is trained on an iterated normal-form game using Q-Learning, likely to converge to an equilibrium as opposed
to displaying complex, or unstable behaviours? In investigating this
question, we will make the following assumptions:

@@

Analyses have been performed from the perspective of the replicator
dynamic w.r.t.~the various types of behaviours that can emerge from
iterated games. Such games are considered imperfect, in that agents
make decisions by attempting to anticipate their opponents' behaviour
based on experience \cite{Galla2011}. \cite{Imhof2005} presents the
observation that, in the Prisoner's Dilemma, cyclic behaviour
emerges in which the optimal strategy cycles across defection,
cooperation and tit-for-tat despite defection being the
NE. In \cite{Galla2011}, Galla builds on this observation by showing that, in fact,
such behaviour is highly dependent on the presence of memory loss in
the system. In the absence of memory loss, the learning converges to
the NE as expected.
%However, as this parameter is increased, the stability of the NE is
%removed and other stable points are created.

Taking the notion of parameter dependence further, \cite{Galla2013}
describes an analysis of two-player iterated games in which the
players learn using \textit{experience weighted attraction} (EWA), a
form of reinforcement learning typically applied in experimental
economics \cite{Camerer2009}. Through numerical simulations, the
authors are able to show that the emergence of chaos and cycles is
dependent on the choice of parameters.
%(see \cite{Strogatz2000} for
%details on chaotic and periodic motion).
A rigorous theoretical analysis of the replicator dynamic
corresponding to EWA, which applies the techniques in
\cite{Opper1992}, results in a method for characterising the regions
in parameter space in which a game is likely to converge, exhibit
chaos or limit cycles. More recently, the same authors presented an
extended study in \cite{Sanders2018}, in which the same analysis is
applied to generic $p$-player games, which showed that chaotic
dynamics are more likely to be observed as the number $p$ of players
increases. Differently from this line, we here consider 
%By contrast to the present work, \cite{Barber2012} focuses
%on the analysis of the EWA algorithm, while we consider
 Q-Learning \cite{Barber2012}, which is one of the most popular
 algorithms in RL.

@@

\section{Preliminaries}

@@

{\small
\begin{equation}
\label{eqn::separatePayoffs}
	\begin{split}
	Z(\Vec{\psi}) = \int D[\Vec{x}, \Vec{\hat{x}}] exp( i \sum_{i, \mu} \int dt [ \hxmu{i} (\frac{\dxmu{i}}{\xmu{i}{\mu}} - \talpha \rho_i^\mu (t) - h_i^\mu (t))]) \times \\ exp(-i \alpha \ttau \sum_{\mu} \sum_{i_\mu, i_{-\mu}} \int dt [\hxmu{i} \payoff{i}{\mu} \prod_{\kappa \neq \mu} \xmu{i}{\kappa} )])) 
    \times \\ exp(-i \alpha \ttau \sum_{\mu} \sum_{j_\mu, i_\mu, i_{-\mu}} \int dt [\hxmu{j}  \xmu{i}{\mu} \payoff{i}{\mu} \prod_{\kappa \neq \mu} \xmu{i}{\kappa}])) 
	\times \\ exp(i \sum_{i, \mu}
	\int dt[\xmu{i} \psi^\mu_i(t)]),
\end{split}
\end{equation}
}
where the `generating fields' \cite{Coolen2005} $\Vec{\psi_i(t)}$ and
$\Vec{\phi_i(t)}$ will be set to zero at the end of the
calculation.

The last two exponentials contain the payoffs of the game. These are randomly generated using a multi-variate gaussian and then held fixed for the rest of the game. As such these comprise the aforementioned `quenched disorder'. We will average over this quenched disorder (and therefore average over all possible realisations of payoff matrices) using the mean and covariance expressions given in (\ref{eqn::Payoffs}). We define $Q$ to be the product of the second and third exponentials in (\ref{eqn::separatePayoffs}) 
%
\begin{equation}
\begin{split}
        \mathbb{E}[Q] = exp(- \frac{\alpha^2 \ttau ^2}{2} N \sum_{\mu} \Big [ \int dt dt' L^\mu(t, t') \prod_{\kappa \neq \mu} C^\kappa (t, t') + \\ \Gamma \sum_{\nu \neq \mu} K^\mu (t, t') K^\nu (t, t') \prod_{\kappa \not\in \{\mu, \nu\}} C^\kappa (t, t') \Big ] ) \times \\
        exp(- \frac{\alpha^2 \htau ^2}{2} N \sum_{\mu} \Big [ \int dt dt' L^\mu(t, t') C^\mu (t, t') \prod_{\kappa \neq \mu} C^\kappa (t, t') + \\ \Gamma \sum_{\nu \neq \mu} A^{\mu \nu} (t, t') C^\nu (t, t') C^\mu (t, t') \prod_{\kappa \not\in \{\mu, \nu\}} C^\kappa (t, t') \Big ] )
\end{split}
\end{equation}
%
in which
%
\begin{eqnarray*}
\label{eqn::correlations}
%    \begin{split}
        C^\mu (t, t') &  := & N^{-1} \sum_i \xmu{i}{\mu} \xmudash{i}{\mu} \\
        L^\mu (t, t') & :=  & N^{-1} \sum_i \hxmu{i} \hxmudash{i} \\
        K^\mu (t, t') & :=  & N^{-1} \sum_i \xmu{i}{\mu} \hxmudash{i} \\
        A^{\mu, \nu} (t, t') & := &  N^{-1} \sum_i \hxmu{i} \hxnudash{i}.
%    \end{split}
\end{eqnarray*}

@@

We then examine the
long-term behaviour of the perturbation $\xpert$. by taking
the Fourier transform of (\ref{eqn::Linearised}) and analysing its
behaviour at $\omega = 0$. After some manipulation (Sec. S4), this gives
%
\begin{equation}
%\begin{split}
        \mathbb{E}[|\mathcal{X}(\omega = 0)|^2] = \frac{1}{\left[ (-\alpha^2 \ttau^2 \Gamma q^{p-2} \chi)^2 - 2 (\alpha \ttau + \alpha \htau)^2 (p-1)q^{p-2}\right]} 
%\end{split}
\end{equation}

@@

We illustrate these implications in Figure \ref{fig:theory} which plots the value obtained by the right hand side of (\ref{eqn::Final}). In order to determine these values, we first had to solve (\ref{eqn::fixed_point}) for $q$ and $\chi$. We determine the $\xfixed$ which solves the expression inside the square brackets of (\ref{eqn::fixed_point}) using a Newton-Raphson root-finding approach and use the following expressions for $q, \chi$ to determine their value. 
%
\begin{eqnarray*}
%    \begin{split}
        1 & = & \frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \xfixed(z) exp(-\frac{z^2}{2}) dz \\ 
        q & = & \frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \xfixed(z)^2 exp(-\frac{z^2}{2}) dz \\ 
        \chi & = & \frac{1}{q^{n/2}}\frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \frac{\partial \xfixed (z)}{\partial z} exp(-\frac{z^2}{2}) dz
%    \end{split}
\end{eqnarray*}

@@

In this section, we describe and discuss the numerical experiments we use to verify the implications described in the discussion of Sec. \ref{sec::Theory}. We first describe how these experiments were conducted and then, having presented the results in Figures \ref{fig:NumericalExperiments} and \ref{fig:pVarianceExpt}, we discuss the correlation with the theory.

@@

To generate the numerical simulations in Figure~\ref{fig:NumericalExperiments} we used the
following procedure.
\begin{enumerate}
   \item Fix the parameters $\tau, \gamma$. The former is held at 0.05 and the latter at 0.1.
   \item Initialise values of $\Gamma, \alpha$ (or $\tau$ appropriately). These will be swept over in the experiment.
\item Generate 15 payoff matrices by sampling from a multi-variate Gaussian 
(variables are the payoff elements) with mean zero and covariance parameterised by $\Gamma$.
\item For each of these payoff matrices, initialise a set of agents with random initial conditions (i.e., random action probabilities).
\item Allow both sets of agents to learn over a maximum of $1.5 \times 10^4$ iterations.
\item Keep track of the action probabilities over a window of 500 iterations. At the end of each window, determine the percentage difference between the maximum and minimum values of each strategy component
\item If the difference is less than 1\% consider the game converged. Otherwise continue to the next window.
\item If the game reaches $1.5 \times 10^4$ iterations without satisfying the relative distance criterion, consider it to be non-convergent. Determine the fraction of these 15 games which have converged.

@@

Finally, Figure \ref{fig:pVarianceExpt} plot the degree variation about the fixed point displayed as the number of players $p$ increases. The variation is determined by first allowing the game to iterate for 5000 steps, so that it can be assumed to have reached the vicinity of the fixed point. Then, the action probabilities are recorded for a further 5000 iterations. At the end of this second period, the variance of the actions are calculated as
%
\begin{equation*}
    V(t) = \frac{1}{N} \sum_i \frac{1}{5000} \sum_t x_i(t)^2 - \left[\frac{1}{5000} \sum_t x_i(t) \right]^2 .
\end{equation*}

@@

% The present work represents a first step towards considering the stability
% of multi-agent reinforcement learning and focuses on the limiting case
% of stateless normal-form games played by homogeneous agents.
As
research into the dynamics of RL algorithms
progresses, it would be prudent to apply this analysis to various
other algorithm. Algorithms whose dynamics are established, and are
therefore open to a stability analysis, include piecewise Q-Learning
and Cross Learning. This would provide a strong method by which to
compare and provide safety guarantees to different algorithms for a
particular use case.

>>>>

\begin{abstract}
Modelling the dynamics of Q-Learning is an active and important topic for the sake of developing an
{\em a priori} understanding of Reinforcement Learning. In this paper, we use methods from
evolutionary game theory to analyse the stability of Q-Learning in $p$-player normal form games,
where payoffs are randomly generated. We determine the parameter range in which Q-Learning is
expected to settle to a stable fixed point and the range in which the dynamics are unstable.

This study allows for parameters to be appropriately chosen to ensure the safe convergence of a
learning algorithm and is a first step towards understanding the range of behaviours that can be
displayed by learning using the Q-Learning algorithm. We validate our theoretical results through
numerical simulation and show that, within the bounds of experimental error, the region of
instability can be characterised by the learning dynamics.
\end{abstract}

@@

Single agent reinforcement learning (RL) is a well-established
framework for allowing agents to learn optimal strategies when trained
on an iterated task
%For a particularly strong example of this see
\cite{Vinyals2019}. However, for the realisation of complex tasks,
such as air traffic control, market negotiations, and multi-robot
coordination, it is required that the system be modelled as a
multi-agent system (MAS). Such systems are not as well-understood as
single-agent RL, since a given agent is tasked with
optimising a reward function which depends not only on a
non-stationary environment, but also on the actions of other, possibly
loosely coupled, agents \cite{SchwartzMulti-agentApproach}.

It is, therefore, of paramount importance to develop a strong
theoretical understanding of multi-agent reinforcement learning (MARL)
to allow an {\em a priori} understanding of the behaviour of a given
learning algorithm. Fortunately, the study of MAS is not unique to the
field of AI and has been extensively investigated from the point of view of
economics and game theory as well
\cite{ShohamMultiagentFoundations}. In particular,
%the study of
evolutionary game theory (EGT) considers the problem of a MAS which
repeatedly plays an iterated game. This idea shares a strong
resemblance with MARL and, in fact, in \cite{Tuyls2006AnGames} it was
shown that techniques from EGT may be fruitfully applied to the analysis of
Q-Learning \cite{Sutton2018}.

@@

\paragraph{Contribution}
In this study, we analyse the stability of the dynamics of Q-Learning established by Tuyls et al.
\cite{Tuyls2006AnGames}, which we discuss in Sec. \ref{sec::Prelims}. From this perspective, We
consider the question: under what choice of parameters is a multi-agent system, which is trained on
an iterated normal-form game using Q-Learning, likely to converge to an equilibrium as opposed to
displaying complex, or unstable behaviours? In investigating this question, we will make the
following assumptions:

@@

Analyses have been performed from the perspective of the replicator
dynamic w.r.t.~the various types of behaviours that can emerge from
iterated games. Such games are considered imperfect, in that agents
make decisions by attempting to anticipate their opponents' behaviour
based on experience \cite{Galla2011}. \cite{Imhof2005}, studies the evolutionary dynamics of
strategies in the iterated Prisoner's Dilemma game, in which cyclic behaviour
emerges. Here, the optimal strategy cycles across defection,
cooperation and tit-for-tat despite defection being the
NE. In \cite{Galla2011}, Galla builds on this observation by showing that, in fact,
such behaviour is highly dependent on the presence of memory loss in
the system. In the absence of memory loss, the learning converges to
the NE as expected.
%However, as this parameter is increased, the stability of the NE is
%removed and other stable points are created.

Taking the notion of property (e.g. memory loss) dependence further, \cite{Galla2013}
describes an analysis of two-player iterated games in which the
players learn using \textit{experience weighted attraction} (EWA), a
form of reinforcement learning typically applied in experimental
economics \cite{Camerer2009}. Through numerical simulations, the
authors are able to show that the emergence of chaos and cycles is
dependent on the choice of parameters.
%(see \cite{Strogatz2000} for
%details on chaotic and periodic motion).
A rigorous theoretical analysis of the replicator dynamic
corresponding to EWA, which applies the techniques in
\cite{Opper1992}, results in a method for characterising the regions
in parameter space in which a game is likely to converge, exhibit
chaos or limit cycles. More recently, the same authors presented an
extended study in \cite{Sanders2018}, in which the same analysis is
applied to generic $p$-player games, which showed that chaotic
dynamics are more likely to be observed as the number $p$ of players
increases. Differently from this line, we here consider 
%By contrast to the present work, \cite{Barber2012} focuses
%on the analysis of the EWA algorithm, while we consider
 Q-Learning \cite{Barber2012}, which is one of the most popular
 algorithms in RL.

@@

\section{Preliminaries} \label{sec::Prelims}

@@

{\small
\begin{equation}
\label{eqn::separatePayoffs}
	\begin{split}
	Z(\Vec{\psi}) = \int D[\Vec{x}, \Vec{\hat{x}}] \exp( i \sum_{i, \mu} \int dt [ \hxmu{i} (\frac{\dxmu{i}}{\xmu{i}{\mu}} - \talpha \rho_i^\mu (t) - h_i^\mu (t))]) \times \\ \exp(-i \alpha \ttau \sum_{\mu} \sum_{i_\mu, i_{-\mu}} \int dt [\hxmu{i} \payoff{i}{\mu} \prod_{\kappa \neq \mu} \xmu{i}{\kappa} )])) 
    \times \\ \exp(-i \alpha \ttau \sum_{\mu} \sum_{j_\mu, i_\mu, i_{-\mu}} \int dt [\hxmu{j}  \xmu{i}{\mu} \payoff{i}{\mu} \prod_{\kappa \neq \mu} \xmu{i}{\kappa}])) 
	\times \\ \exp(i \sum_{i, \mu}
	\int dt[\xmu{i} \psi^\mu_i(t)]),
\end{split}
\end{equation}
}
where the `generating fields' \cite{Coolen2005} $\Vec{\psi_i(t)}$ and
$\Vec{\phi_i(t)}$ will be set to zero at the end of the
calculation.

The last two exponentials contain the payoffs of the game. These are randomly generated using a multi-variate gaussian and then held fixed for the rest of the game. As such these comprise the aforementioned `quenched disorder'. We will average over this quenched disorder (and therefore average over all possible realisations of payoff matrices) using the mean and covariance expressions given in (\ref{eqn::Payoffs}). We define $Q$ to be the product of the second and third exponentials in (\ref{eqn::separatePayoffs}) 
%
\begin{equation}
\begin{split}
        \mathbb{E}[Q] = \exp(- \frac{\alpha^2 \ttau ^2}{2} N \sum_{\mu} \Big [ \int dt dt' L^\mu(t, t') \prod_{\kappa \neq \mu} C^\kappa (t, t') + \\ \Gamma \sum_{\nu \neq \mu} K^\mu (t, t') K^\nu (t, t') \prod_{\kappa \not\in \{\mu, \nu\}} C^\kappa (t, t') \Big ] ) \times \\
        \exp(- \frac{\alpha^2 \htau ^2}{2} N \sum_{\mu} \Big [ \int dt dt' L^\mu(t, t') C^\mu (t, t') \prod_{\kappa \neq \mu} C^\kappa (t, t') + \\ \Gamma \sum_{\nu \neq \mu} A^{\mu \nu} (t, t') C^\nu (t, t') C^\mu (t, t') \prod_{\kappa \not\in \{\mu, \nu\}} C^\kappa (t, t') \Big ] )
\end{split}
\end{equation}
%
in which
%
\begin{eqnarray*}
\label{eqn::correlations}
%    \begin{split}
        C^\mu (t, t') &  := & N^{-1} \sum_i \xmu{i}{\mu} \xmudash{i}{\mu} \\
        L^\mu (t, t') & :=  & N^{-1} \sum_i \hxmu{i} \hxmudash{i} \\
        K^\mu (t, t') & :=  & N^{-1} \sum_i \xmu{i}{\mu} \hxmudash{i} \\
        A^{\mu, \nu} (t, t') & := &  N^{-1} \sum_i \hxmu{i} \hxnudash{i}.
%    \end{split}
\end{eqnarray*}

@@

We then examine the
long-term behaviour of the perturbation $\xpert$ by taking
the Fourier transform of (\ref{eqn::Linearised}) and analysing its
behaviour at $\omega = 0$. After some manipulation (Sec. S4), this gives
%
\begin{equation}
%\begin{split}
        \mathbb{E}[|\mathcal{X}(\omega = 0)|^2] = \frac{1}{\left[ (-\alpha^2 \ttau^2 \Gamma q^{p-2} \chi)^2 - 2 (\alpha \ttau + \alpha \htau)^2 (p-1)q^{p-2}\right]} 
%\end{split}
\end{equation}

@@

We illustrate these implications in Figure \ref{fig:theory} which plots the value obtained by the right hand side of (\ref{eqn::Final}). In order to determine these values, we first had to solve (\ref{eqn::fixed_point}) for $q$ and $\chi$. We determine the $\xfixed$ which solves the expression inside the square brackets of (\ref{eqn::fixed_point}) using a Newton-Raphson root-finding approach and use the following expressions for $q, \chi$ to determine their value. 
%
\begin{eqnarray*}
%    \begin{split}
        1 & = & \frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \xfixed(z) \exp(-\frac{z^2}{2}) dz \\ 
        q & = & \frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \xfixed(z)^2 \exp(-\frac{z^2}{2}) dz \\ 
        \chi & = & \frac{1}{q^{n/2}}\frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \frac{\partial \xfixed (z)}{\partial z} \exp(-\frac{z^2}{2}) dz
%    \end{split}
\end{eqnarray*}

@@

In this section, we describe and discuss the numerical experiments we use to verify the implications
described in the discussion of Sec. \ref{sec::Theory}. We first describe how these experiments were
conducted and then, having presented the results in Figures \ref{fig:NumericalExperiments} and 
\ref{fig:pVarianceExpt}, we discuss the correlation with the theory. Of course, (\ref{eqn::Final}) 
holds only in the limit of a large player set and $N \rightarrow \infty$, which is impossible to
achieve experimentally. Our aim, therefore, is to show that, by assessing the stability at
different choices of $p$, that the trend is towards everywhere instability across parameter
space. Specifically, as $p$ increases, the likelihood of instability increases, regardless of
parameter choice.

@@

To generate the numerical simulations in Figure~\ref{fig:NumericalExperiments} we used the
following procedure.
\begin{enumerate}
   \item Fix the parameters $\tau, \gamma$. The former is held at 0.05 (to remain consistent with
   \cite{Sanders2018}) and the latter at 0.1.
   \item Initialise values of $\Gamma, \alpha$ (or $\tau$ appropriately). These will be swept over in the experiment.
\item Generate 15 payoff matrices by sampling from a multi-variate Gaussian 
(variables are the payoff elements) with mean zero and covariance parameterised by $\Gamma$.
\item For each of these payoff matrices, initialise a set of agents with random initial conditions (i.e., random action probabilities).
\item Allow both sets of agents to learn over a maximum of $1.5 \times 10^4$ iterations.
\item Keep track of the action probabilities over a window of 500 iterations. At the end of each window, determine the percentage difference between the maximum and minimum values of each strategy component
\item If the difference is less than 1\% consider the game converged. Otherwise continue to the next window.
\item If the game reaches $1.5 \times 10^4$ iterations without satisfying the relative distance criterion, consider it to be non-convergent. Determine the fraction of these 15 games which have converged.

@@

Finally, Figure \ref{fig:pVarianceExpt} plot the degree variation about the fixed point displayed as
the number of players $p$ increases. The variation is determined by first allowing the game to
iterate for 5000 steps (times steps are denoted with $t$), so that it can be assumed to have reached
the vicinity of the fixed point. Then, the action probabilities $\x_i$ are recorded for a further
5000 iterations. At the end of this second period, the variance of the actions are calculated as
%
\begin{equation*}
    V(t) = \frac{1}{N} \sum_i \frac{1}{5000} \sum_t x_i(t)^2 - \left[\frac{1}{5000} \sum_t x_i(t) \right]^2 .
\end{equation*}

@@

% The present work represents a first step towards considering the stability
% of multi-agent reinforcement learning and focuses on the limiting case
% of stateless normal-form games played by homogeneous agents.
As
research into the dynamics of RL algorithms
progresses, it would be prudent to apply this analysis to various
other algorithms. Algorithms whose dynamics are established, and are
therefore open to a stability analysis, include piecewise Q-Learning
and Cross Learning. This would provide a strong method by which to
compare and provide safety guarantees to different algorithms for a
particular use case.
